Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-09-01 paper/letter.tex
comments from mark

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.   
Recent advances in these computational models have explored the
effects that bulk rotation and a multiphase state in the interstellar
medium should have on the Lyman-$\alpha$ line.

The most important consequence of bulk rotation is that, even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is high enough, the line becomes single
peaked and symmetric, a unique feature that other theoretical models
find difficult to reproduce without introducing more complexity into
the gas distribution.

This is the case of a multiphase interstellar medium.
The model is more complex as it must take into account the behavior
of neutral dense hydrogen clumps dispersed in a tenuous intraclump medium.  
Corresponding to the elaborate physical conditions in the model,
there are a wide variety of possible emerging spectra.
This variety includes single, double and triple peaked
spectra with different degrees of symmetry around the line's center.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ had evaded a physical
interpretation so far. 
Including the effects of either bulk rotation or a multiphase ISM, we
manage to explain the \lya\ line in \tol.

Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continuous line) and a clumpy ISM
(dashed line).

>>>>

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ cannot be explained with
conventional models.

In this letter we show how the \tol's \lya\ profile can be explained
either by rotation or the recently developed class of more complex
multiphase models. These models predict a wider variety of spectra
including, single, double and triply peaked spectra.  
Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fit models from the analytical solution for a
rotating homogeneous gas sphere (continuous line) and a clumpy ISM
(dashed line).
This is the first time that these models have been introduced with
success to explain the a \lya\ profile.
